Water discharge has a variable impact on sediment transport capacity in subglacial systems

### 2023-07-10 Delaney2023.tex
more stumbling

\title{Comparing variations in sediment transport capacity between subglacial and subaerial channels }

@@

\begin{keypoints}
\item Water discharge variations in subglacial channels are accommodated by water velocity, not channel size, given their slow evolution.

@@

\begin{abstract} % 150 words
Sediment transport capacity depends mainly on the shear stress or velocity of water flowing through a channel and the channel width over which to mobilize sediment.
In subaerial channels, changing water discharge can be accommodated by rapid changes in the channel width over which water flows, the water depth, and the water velocity.
In subglacial channels, however, water flows through a conduit whose size evolves  slowly, because the water is pressurized by the glacier ice above.
As a result, variations in water discharge may modify the water velocity, but not the channel's width.
Here, parameterizations of subglacial and subaerial channels indicate that sediment transport capacity varies more in subglacial systems compared with subaerial systems, across a wide range of channel geometries and flow conditions.
Large variations in subglacial water sediment transport capacity may impact sediment transport processes and the interpretation of records of sediment transport from glacierized catchments.

@@

\section*{Plain language summary} % 200 words
In both subglacial and subaerial channels, sediment transport capacity depends mainly on 1) the channel width over which the sediment is mobilized  and 2) the shear stress applied by flowing water on sediment at the channel bottom; the shear stress depends mainly on the water velocity.
In subaerial channels, changing water discharge will  change the water depth,  the channel width (which dictates the distance  over which sediment transport  occurs), and the  water velocity (which controls the sediment transport capacity).
 In subglacial environments, however, on times scales of hours to weeks, changing water discharge will only alter  water velocity, not channel area, because the subglacial conduit maintains a largely fixed geometry.
The fixed conduit geometry also means that the channel width over which the sediment is  mobilized also does not respond to changing water discharge.
Parameterizations of these processes show that sediment transport capacity varies far more in subglacial systems compared to subaerial systems, even though changing channel width can accommodate some sediment transport variability in subaerial channels.
The increased variability in sediment transport capacity in subglacial systems, compared to subaerial ones, has several important implications for sediment transport in glacierized catchments and the interpretation of sediment discharge records.

@@

%%% Suggested section heads:
\section{Introduction}
\label{sect:intro}

  Glaciers expel large quantities of sediment from their termini \cite{hallet1996}.
  Changes in glacier dynamics, geomorphology, and hydrology  have prompted numerous  recent studies of  sediment transport processes in cold regions \cite<e.g.>[]{zhang2022}.
  Increases in sediment transport  have been observed in Greenland \cite{bendixen2017}, the European Alps \cite{costa2017}, the Himalayas \cite{li2021}, and the Andes \cite{vergara2022}.
  In some of these regions, increased water discharge and glacier melt have been attributed to greater sediment transport \cite{bendixen2017,costa2017,li2021}.
  In other cases, enhanced access of meltwater to subglacial sediment at high elevations was attributed as the cause of increased sediment transport \cite{delaney2020,vergara2022}.
  Observed changes to sediment transport in glacierized catchments create an imperative to examine the processes controlling sediment discharge here. 
  Yet, the relationship between water discharge and sediment transport in subglacial environments remains minimally discussed and not contrasted with the subaerial river systems that these glaciers feed.

@@

  In a transport-limited regime, sediment discharge responds to the sediment transport capacity or the amount of sediment that could be carried by the water. Sediment transport capacity depends on the shear stress between water and sediment it flows  over \cite{shields1936,meyer1948,engelund1967} and the width of the channel bottom $w_c$ over which to mobilize sediment. The shear stress  across the channel width $\tau_t$ responds to the velocity of water $v$ flowing through at channel so that 
  \begin{linenomath*}
    \begin{equation}
      \label{eq:tau_t}
      \tau_t \propto w_c\, v^2,
    \end{equation}
  \end{linenomath*}
  % 
  where $w_c$ is the width of the channel.
  Following the conservation of mass, the velocity of the water flowing through a rectangular channel is given as 
  \begin{linenomath*}
    \begin{equation}
      \label{eq:v}
      v = \frac{Q_w}{S},
    \end{equation}
  \end{linenomath*}
  % 
  where $Q_w$ is water discharge,  and $S$ is the channel's cross-sectional area.

  In subaerial channels, with open channel flow, the cross-sectional area $S$ of the channel evolves with changing water discharge $Q_w$  by changing the width of the channel along with the depth \cite{leopold1953}. During a flood with greater water discharge $Q_w$, for instance, the cross-sectional area $S$ increases as the channel widens and deepens, along with water velocity $v$. As a result, the shear stress across the channel $\tau_t$ will increase in response to both the channel width $w_c$  and water velocity $v$.

  Yet, the response of water velocity to changing water discharge in subglacial channels is different than subaerial ones.
  In subglacial systems, the water is pressurized by the ice above \cite{shreve1972}.
  The size of subglacial channels responds to the creep closure of the ice above and the opening of the channel by frictional heating of water flowing through the channel \cite{rothlisberger1972}.
  As a result, the subglacial channel size evolves over days to months, whereas water discharge varies over hours \cite<e.g.>[]{iken1986,andrews2014,nanni2020}.
  In turn, changes in water discharge $Q_w$ are mainly accommodated with water velocity $v$, because the size of the channel $S$ is largely fixed (Equation~\ref{eq:v} and Figure~\ref{fig:cartoon}).
  It follows that greater water discharge $Q_w$ will cause the shear stress across the channel $\tau_t$ to respond to the water velocity $v$, as opposed to the subaerial system where width evolves as well.

  As a result, sediment mobilization in subaerial and subglacial systems responds differently to changing water discharge.
  These differences have been implicitly included in a wide range of available models that  quantify sediment transport from catchments in both subglacial and subaerial systems  \cite<e.g.>[]{walder1994,tucker1997,creyts2013,wickert2019,hewitt2019}.
  The divergent response of changing water discharge on sediment mobilization capacity may well impact sediment dynamics at glacier margins where flow transitions from pressurized to open channel flow \cite<e.g.>[]{lane2016}.
  Additionally, the variable response of sediment transport capacity to water discharge in the two systems may impact the interpretation of records of sediment transport in glacial compared to subaerial systems \cite<e.g.>[]{muller1968,richards2003,ganti2016}.
  Yet, the differing relationship between variations in water discharge and sediment discharge in subglacial and subaerial systems has been minimally discussed.

  This manuscript leverages parameterizations of subglacial and subaerial hydraulics to explore differences in the relationships between  water discharge, channel geometry, and water velocity. 
  With these parameterizations, the manuscript examines the impact of water discharge variability on sediment transport capacity in idealized subglacial and subaerial systems.
  Lastly, the manuscript discusses the implications for sediment transport processes in subglacial systems, compared to subaerial ones.

  \begin{center}
    \begin{figure}[H]
      \includegraphics[width=0.8\linewidth]{Fig1.pdf}
      \caption{Cartoon of the situation with various parameters. Water velocity magnitude in the subglacial, $v_g$, and subaerial, $v_f$, systems are shown by arrow length. $S_g$ and $S_f$ represent the cross-sectional area in subglacial and subaerial channels. The subglacial channel widths $w_{cg}$ remain steady, while the subaerial channel widths evolve $w_{cf}$, denoted by the thick black line. Shear stress $\tau$ is responsible for the mobilization of sediment and increases with the number of makers at the channel bed. 
        Note that in the subaerial channel parameterization, rectangular channel shapes are implemented (Section~\ref{sect:fluv}).} 
      \label{fig:cartoon}
    \end{figure}
  \end{center}

@@

  \begin{table}[H]
    \centering
    \caption{Variables, parameters, and constants }
    \begin{tabular}{ l  c  c c }
      Name &Symbol&  Value&Units \\ \hline
      \textbf{Variables}  & & & \\
      Channel hydraulic diameter (glacier, subaerial) &  $D_{hg},\,D_{hf}$&  & $\mathrm{m}$     \\
      Width of channel floor (glacial, subaerial) & $w_{cg},w_{cf}$&  & $\mathrm{m}$     \\
      Channel cross-sectional area (glacier, subaerial) &  $S_g, S_f$& & $\mathrm{m^2}$     \\
      Water discharge (instantaneous) & $Q_w$& & $\mathrm{m^{3}\,s^{-1}}$ \\
      Representative water discharge & $Q_{w}^*$& & $\mathrm{m^{3}\,s^{-1}}$ \\
      Representative hydraulic gradient  &$\Psi^*$ & & $\mathrm{Pa\, m^{-1}}$\\
      Water velocity (glacier, subaerial)  & $v_g,\,v_{f}$& & $\mathrm{m\,s^{-1}}$ \\
      Shear stress (glacier, subaerial) & $\tau_g,\,\tau_f$&& $\mathrm{Pa \, m^{-1}}$ \\
      Width-integrated shear-stress (glacier, subaerial) & $\tau_{tg},\, \tau_{tf}$&& $\mathrm{Pa \, m^{-1}}$ \\
      Reynolds number &$R_e$& & $\mathrm{(-)}$\\
      Variable &$\Gamma$&&$\mathrm{(-)}$\\
      Variability of a variable ($\Gamma$) &$\chi$& &$\mathrm{(-)}$\\
           &&&\\

      \textbf{Parameters and Constants}  & & &\\
      Gravitational constant&$g$& $-9.81$&$\mathrm{m\,s^{-2}}$\\
      Density of water & $\rho_w$& $1000$ & $\mathrm{kg\,m^{-3}}$ \\
      Density of ice & $\rho_i$& $900$ & $\mathrm{kg\,m^{-3}}$ \\
      Kinematic viscosity of water &$\nu$& $10^{-4}$& $\mathrm{m^2\,s^{-1}}$\\
      Hooke angle of channel & $\beta$ & $\frac{\pi}{2}$ & \unit{rad}\\
      Flotation fraction & $f_f$&$0.7$& $\mathrm{(-)}$\\
      Friction factor (glacier) & $f_r$ & $4$ & $\mathrm{(-)}$ \\
      Friction factor (subaerial) & $f_p$ & $3$ & $\mathrm{(-)}$\\
      Gradient of glacier surface & $\frac{\partial z_{sg}}{\partial x}$ &$0.25$& $\mathrm{(-)}$\\
      Gradient of channel bed (subaerial) &$\frac{\partial z_c}{\partial x}$ &$0.05$& $\mathrm{(-)}$\\
      Subaerial channel factor & $k$ &$3$ & $\mathrm{s\,m^{-2}}$\\
      Channel geometry exponent &$e$& $\frac{1}{2}$&$\mathrm{(-)}$ \\
      \hline
    \end{tabular}
    \label{table:vpm}
  \end{table}

  \subsection{Subglacial channel  parameterization}
  \label{sect:sub_mode}
  To evaluate the shear stress of water flowing across sediments below a glacier, the subglacial channel parameterization evaluates the channel geometry and the velocity of the flowing water.
  To accomplish this, the hydraulics model presented in \citeA{delaney2019} is used.

  Here, it is assumed that the water is transported through subglacial channels \cite<Figure~\ref{fig:cartoon}; >{rothlisberger1972}, and that the channel  geometry will respond to a representative water discharge over a certain time period $Q_{w}^*$, using the Darcy-Weisbach formulation for water-flow through a pipe  \cite<e.g.>[]{rothlisberger1972,clarke2003,werder2013}.
  Thus, the hydraulic diameter of the subglacial channel, $D_h$, responds to a representative water discharge, $Q_{w}^*$, and a representative gradient of the hydraulic potential, $\Psi^*$ such that:
  \begin{linenomath*}
    \begin{equation}
      \label{eq:DW}
      D_h = \big(s\, f_r\,\rho_w\, \frac{Q_w^{*\,2}}{\Psi^*}\big)^{\frac{1}{5}}~~.
    \end{equation}
  \end{linenomath*}
  % 
  The representative hydraulic gradient is based upon the pressurized flow in Equation~\ref{eq:DW},
  \begin{equation}
    \label{eq:psi}
    \Psi^*= f_f \,  \rho_i \, g (\frac{\partial  z_{sg}}{\partial x} - \frac{\partial z_c}{\partial x}) +  \rho_i \, g \, \frac{\partial z_c}{\partial x},
  \end{equation}
  % 
  \noindent
  where, $f_f$ is the flotation fraction, $\rho_i$ is density of ice $\frac{\partial z_{sg}}{\partial x}$ is the surface slope of the glacier and $\frac{\partial z_c}{\partial x}$ is the glacier bed slope.

  In Equation~\ref{eq:DW}, $f_r$ is the Darcy-Weisbach friction factor and $\rho_w$ is the density of water (Table \ref{table:vpm}). $s$ is a factor accounting for channel geometry \cite{hooke1990}, calculated as:
  \begin{equation}
    \label{eq:Hf}
    s = \frac{2\,(\beta -\sin \beta)^2}{(\frac{\beta}{2}\,+\,\sin \frac{\beta}{2})^4},
  \end{equation}
  where $\beta$ is the central angle of the circular segment that comprises the channel (the so-called Hooke angle). Note that $\beta =\pi$ corresponds to a semi-circle and
  smaller values of $\beta$ result in shallow, wide channels. 
  The width of the flat channel floor $w_c$, given angle $beta$ is 
  \begin{equation}
    \label{eq:dh2wc}
    w_{cg} = 2  \sin \frac{\beta}{2} \sqrt{\frac{2\, S}{\beta -\sin \beta}},
  \end{equation}
  % 
  where $S$ is the cross-sectional area of the channel given by
  \begin{equation}
    \label{eq:dh2S}
    S_g =  \frac{D_h^2}{2}~ \frac{\Big(\frac{\beta}{2} \,+ \, \sin \frac{\beta}{2}\Big)^2  }{\beta\,-\,\sin \beta}.
  \end{equation}

  The shear stress, $\tau$, between the water and the channel bed is determined through the Darcy-Weisbach formulation
  \begin{equation}
    \label{eq:tau}
    \tau_g=\frac{1}{8}\,f_r\,\rho_w\,v_g^2,
  \end{equation}
  % 
  where  the water velocity $v_g$ is $v_g = \frac{Q_w}{S_g}$.
  Here, $Q_w$ represents the instantaneous water discharge, rather than $Q_w^*$, which is used to evaluate the channel geometry.
  With this formulation, the processes in an R-channel, through which water flows subglacially, can be represented algebraically \cite{rothlisberger1972,delaney2019}.

@@

  To parameterize the shear stress of water flowing across sediments in the subaerial channel,  the hydraulics parameterization presented in \citeA{tucker1997} is implemented.
  Here, it is assumed that there is a conservation of mass and a sufficiently wide channel so that the hydraulic radius is consistent with flow depth, uniform flow, and the Darcy-Weisbach relationship.
  The shear stress $\tau_f$ at the river bed is represented as
  \begin{linenomath*}
    \begin{equation}
      \label{eq:DW_tau}
      \tau_f=\frac{\rho_w\,g^{\frac{2}{3}}\,f_p^{\frac{1}{3}}}{2}\, \Big(\frac{Q_w}{w_{cf}} \Big)^{\frac{2}{3}} \,\frac{\partial z_c}{\partial x}^{\frac{2}{3}},
    \end{equation}
  \end{linenomath*}
  where $\frac{\partial z_c}{\partial x}$ is the channel slope, and $f_p$ is the friction factor for subaerial systems.
  Channel width $w_{cf}$ is 
  \begin{equation}
    \label{eq:wcf}
    w_{cf} = k \, Q_w^e,
  \end{equation}
  % 
  where $k$ is a constant and $e$ is an exponent commonly equal to $\frac{1}{2}$ \cite{leopold1953}.
  Cross-sectional area, $ S_f$, is 
  \begin{equation}
    \label{eq:Sf}
    S_f = \frac{Q_w}{v_f},
  \end{equation}
  % 
  where $v_f$ is water velocity, given by rearranging Equation \ref{eq:tau} as
  \begin{equation}
    \label{eq:vf}
    v_f = \sqrt{\frac{8\,\tau_f}{f_p\,\rho_w}}.
  \end{equation}
  % 
  As in Section~\ref{sect:sub_mode}, the width integrated shear stress is $\tau_{tf}=w_{cf}\,\tau_f$.
  Hydraulic diameter of the channel is given by $D_{hf} = \frac{4\,w_{cf}\,d}{2\,d+w_{cf}}$, with flow depth $d$ determined from knowledge of $S_f$ and $w_{cf}$ in a rectangular channel.

@@

  The formulation of the glacial system (Section~\ref{sect:sub_mode}) requires inputs of the hydraulic gradient (Equation \ref{eq:psi}; $\frac{\partial z_{sg}}{\partial x}$, $\frac{\partial z_{c}}{\partial x}$), water discharge, $Q_w$ representative water discharge, $Q_w^*$,  Hooke angle (Equation~\ref{eq:Hf}; $\beta$), and friction factor (Equation~\ref{eq:DW}; $f_r$).

@@

  The Reynolds number $R_e$, which gives an evaluation of the degree of turbulence of the water, is given as 
  \begin{equation}
    \label{eq:re}
    R_e\,=\, v \,\frac{D_h}{\nu},
  \end{equation}
  % 
  \noindent where $\nu$ is the kinematic viscosity of water, and  $v$ is velocity of water, $v_g$ or $v_f$.

  The variance $ \chi$ of a variable $\Gamma$ is given by 
  \begin{equation}
    \label{eq:var}
    \chi(\Gamma) \,=\, 1 - \frac{\mathrm{min}(\Gamma)}{\mathrm{max}(\Gamma)}\,\,.
  \end{equation}
  % 
  \noindent Below we examine variables, $\Gamma$, as water discharge ($Q_w$) water velocity ($v_g$, $v_f$), shear stress ($\tau_g$, $\tau_f$), and integrated shear stress ($\tau_{tg}$, $\tau_{tf}$) with the purpose of evaluating total shear stress across the glacier bed.

>>>>

\title{Water discharge has a variable impact on sediment transport capacity in subglacial systems}

\ian{Alpine glacier vs Ice sheet}

@@

\begin{keypoints}
\item Water discharge variations in subglacial channels are mainly accommodated by water velocity, not channel size, given their slow evolution.

@@

\begin{abstract} % 150 words

In both subglacial and subaerial channels, sediment transport capacity of a sediment size depends mainly on 1) the channel width over which the sediment is mobilized  and 2) the shear stress applied by flowing water on sediment at the channel bottom; the shear stress depends mainly on the water velocity.
In subaerial channels, changing water discharge will  change the water depth,  the channel width (which dictates the distance  over which sediment transport  occurs), and the  water velocity (which controls the sediment transport capacity).
In subglacial environments, however, on times scales of hours to weeks, changing water discharge will only alter  water velocity, not channel area, because the subglacial conduit maintains a largely fixed geometry.
The fixed conduit geometry also means that the channel width over which the sediment is  mobilized also does not respond to changing water discharge.
Parameterizations of these processes show that sediment transport capacity varies far more in subglacial systems compared to subaerial systems, even though changing channel width can accommodate some sediment transport variability in subaerial channels.
The increased variability in sediment transport capacity in subglacial systems, compared to subaerial ones, has several important implications for sediment transport processes in glacierized catchments and the interpretation of sediment discharge records.

@@

@@

%%% Suggested section heads:
\ian{comparison with Alley 1997... mention his variability-- also deal with his $Qs \propto Qw^{\frac{9}{2}}$  treats like rivers?  does not consider channel size evolution}

\section{Introduction}
\label{sect:intro}

Glaciers expel large quantities of sediment from their termini \cite{hallet1996}.
Changes in glacier dynamics, geomorphology, and hydrology  have prompted numerous  recent studies of  sediment transport processes in cold regions \cite<e.g.>[]{zhang2022}.
Increases in sediment transport  have been observed in Greenland \cite{bendixen2017}, the European Alps \cite{costa2017}, the Himalayas \cite{li2021}, and the Andes \cite{vergara2022}.
In some of these regions, increased water discharge and glacier melt have been attributed to greater sediment transport capacity \cite{bendixen2017,costa2017,li2021}.
In other cases, enhanced access of meltwater to subglacial sediment at high elevations was attributed as the cause of increased sediment transport \cite{delaney2020,vergara2022}.
Observed changes to sediment transport in glacierized catchments create an imperative to examine the processes controlling sediment discharge here. 
Yet, the relationship between water discharge and sediment transport in subglacial environments remains minimally discussed and not contrasted with the subaerial river systems that these glaciers feed.

@@

In a transport-limited regime, sediment discharge responds to the sediment transport capacity or the amount of sediment that could be carried by the water. Sediment transport capacity depends on the shear stress between water and sediment it flows  over \cite{shields1936,meyer1948,engelund1967} and the width of the channel bottom $w_c$ over which to mobilize sediment. The shear stress  across the channel width $\tau_t$ responds to the velocity of water $v$ flowing through at channel so that 
\begin{linenomath*}
  \begin{equation}
    \label{eq:tau_t}
    \tau_t \propto w_c\, v^2,
  \end{equation}
\end{linenomath*}
% 
where $w_c$ is the width of the channel.
Following the conservation of mass, the velocity of the water flowing through a  channel is given as 
\begin{linenomath*}
  \begin{equation}
    \label{eq:v}
    v = \frac{Q_w}{S},
  \end{equation}
\end{linenomath*}
% 
where $Q_w$ is water discharge,  and $S$ is the channel's cross-sectional area.

In subaerial channels, with open channel flow, the cross-sectional area $S$ of the channel evolves with changing water discharge $Q_w$  by changing the width of the channel along with the depth \cite{leopold1953}. During a flood with greater water discharge $Q_w$, for instance, the cross-sectional area $S$ increases as the channel widens and deepens, along with water velocity $v$. As a result, the shear stress across the channel $\tau_t$ will increase in response to both the channel width $w_c$  and water velocity $v$, which is squared.

Yet, the response of water velocity to changing water discharge in subglacial channels is different than subaerial ones.
In subglacial systems, the water is pressurized by the ice above \cite{rothlisberger1972}.
The size of subglacial channels responds to the creep closure of the ice above and the opening of the channel by frictional heating of water flowing through the channel \cite{rothlisberger1972}.
As a result, the subglacial channel size evolves over days to months, whereas water discharge can vary over hours \cite<e.g.>[]{iken1986,andrews2014,nanni2020}.
In turn, changes in water discharge $Q_w$ are mainly accommodated with water velocity $v$, because the size of the channel $S$ changes on a much shorter timescale (Equation~\ref{eq:v} and Figure~\ref{fig:cartoon}).
It follows that an increase in water discharge $Q_w$ will cause the shear stress across the channel $\tau_t$ to respond to the water velocity $v$, as opposed to the subaerial system where channel width $w_c$ evolves as well.

As a result, sediment mobilization in subaerial and subglacial systems responds differently to changing water discharge.
These differences have been implicitly included in a wide range of available models that  quantify sediment transport from catchments in both subglacial and subaerial systems  \cite<e.g.>[]{walder1994,tucker1997,creyts2013,wickert2019,hewitt2019}.
The divergent response of changing water discharge on sediment mobilization capacity may well impact sediment dynamics at glacier margins where flow transitions from pressurized to open channel flow \cite<e.g.>[]{lane2016,perolo2018}.
Additionally, the variable response of sediment transport capacity to water discharge in the two systems may impact the interpretation of records of sediment transport in glacial compared to subaerial systems \cite<e.g.>[]{muller1968,richards2003,ganti2016}.
Yet, the differing relationship between variations in water discharge and sediment discharge in subglacial and subaerial systems has been minimally discussed.
Furthermore, thorough observations of water velocity and channel width in subglacial systems are limited and not contrasted with subaerial conditions at the glacier margins.

To better understand the relationship between sediment transport capacity and water discharge in subglacial systems, this manuscript leverages parameterizations of subglacial and subaerial hydraulics.
These parameterizations are applied to hydrological records from the Fieschergletscher, an alpine glacier in Switzerland, and Leverett glacier,  a land-terminating glacier in Greenland.
Outputs are used to explore differences in the relationships between  water discharge, channel geometry, and water velocity in the subglacial and subaerial. 
With these parameterizations, the manuscript examines the impact of water discharge variability on sediment transport capacity in these subglacial and subaerial systems.
The manuscript then uses the demonstrated differences between the two systems to discuss the implications for sediment transport processes in subglacial systems, compared to subaerial ones, and for interpreting records of sediment transport from glacierized catchments.

\begin{center}
  \begin{figure}[H]
    \includegraphics[width=0.8\linewidth]{Fig1.pdf}
    \caption{Sketch of the situation with various parameters. Water velocity magnitude in the subglacial, $v_g$, and subaerial, $v_f$, systems are shown by arrow length. $S_g$ and $S_f$ represent the cross-sectional area in subglacial and subaerial channels. The subglacial channel widths $w_{cg}$ remain steady, while the subaerial channel widths evolve $w_{cf}$, denoted by the thick black line. Shear stress $\tau$ is responsible for the mobilization of sediment and increases with the number of makers at the channel bed. 
      Note that in the subaerial channel parameterization, rectangular channel shapes are implemented (Section~\ref{sect:fluv}).} 
    \label{fig:cartoon}
  \end{figure}
\end{center}

@@

\ian{update hooke angle }
\begin{table}[H]
  \centering
  \caption{Variables, parameters, and constants }
  \begin{tabular}{ l  c  c c }
    Name &Symbol&  Value&Units \\ \hline
    \textbf{Variables}  & & & \\
    Water discharge  & $Q_w$& & $\mathrm{m^{3}\,s^{-1}}$ \\
    Water velocity (glacier, subaerial)  & $v_g,\,v_{f}$& & $\mathrm{m\,s^{-1}}$ \\
    Channel cross-sectional area (glacier, subaerial) &  $S_g, S_f$& & $\mathrm{m^2}$     \\
    Hydraulic diameter &$D_h$&&$\mathrm{m}$\\
    Width of channel floor (glacial, subaerial) & $w_{cg},w_{cf}$&  & $\mathrm{m}$     \\
    Hydraulic head &$\Delta h$&& $\mathrm{m}$\\
    Shear stress (glacier, subaerial) & $\tau_g,\,\tau_f$&& $\mathrm{Pa \, m^{-1}}$ \\
    Width-integrated shear-stress (glacier, subaerial) & $\tau_{tg},\, \tau_{tf}$&& $\mathrm{Pa \, m^{-1}}$ \\

         &&&\\

    \textbf{Parameters and Constants}  & & &\\
    Gravitational constant&$g$& $-9.81$&$\mathrm{m\,s^{-2}}$\\
    Density of water & $\rho_w$& $1000$ & $\mathrm{kg\,m^{-3}}$ \\
    Density of ice & $\rho_i$& $900$ & $\mathrm{kg\,m^{-3}}$ \\
    Hooke angle of channel & $\beta$ & $\frac{\pi}{2}$ & \unit{rad}\\
    Effective glacier thickness &$h_o$& $225^a$ or $700^b$  &\unit{m}\\
    Effective glacier length &$l$&$7000^a$ or $20,000^b$&\unit{m}\\
    Constant $1$ &$C_1$&$2.2\times10^{-5}$&\unit{m}$^{-1}$\\
    Constant $2$ &$C_2$&$3.7\times10^{-13}$&\unit{m}$^{-n}\,s^{-1}$\\
    Latent heat of fusion &$L$&$333.5 $&\unit{kJ\,kg}$^{-1}$\\
    Pressure melting coefficient &$c_t$&$7.5\times 10^{-8}$&\unit{K\,Pa}$^{-1}$\\
    Specific heat capacity of water &$c_p$&$4180$&\unit{J\,kg}$^{-1}$\unit{K}$^{-1}$\\

    Ice flow constant &$A$& $5.3\times10^{-24}$ &\unit{Pa}$^{-n}$\,$s^{-1}$\\
    Ice flow exponent &$n$& $3$ &$\mathrm{(-)}$\\
    Friction factor (glacier) & $f_r$ & $4$ & $\mathrm{(-)}$ \\
    Friction factor (subaerial) & $f_p$ & $3$ & $\mathrm{(-)}$\\
    Gradient of channel bed (subaerial) &$\frac{\partial z_c}{\partial x}$ &$0.05$& $\mathrm{(-)}$\\
    Subaerial channel factor & $k$ &$3$ & $\mathrm{s\,m^{-2}}$\\
    Channel geometry exponent &$e$& $\frac{1}{2}$&$\mathrm{(-)}$ \\
    \hline
  \end{tabular}
  \label{table:vpm}

\end{table}
  $^a$ for Fieschergletscher

  $^b$ for Leverett glacier

\subsection{Subglacial channel  parameterization}
\label{sect:sub_mode}

To evaluate the shear stress of water flowing across sediments below a glacier, the subglacial channel parameterization evaluates the channel geometry and the velocity of the flowing water.
To accomplish this, a modified version of  the lumped hydraulics model presented in \citeA{clarke1996} and \citeA{werder2010} is used.

Here, it is assumed that the water is transported through a subglacial channel flowing down the glacier; \cite<Figure~\ref{fig:cartoon}; >{rothlisberger1972}, and that the channel  size responds to frictional heating from water flow and creep closure by ice, using the Darcy-Weisbach formulation for water-flow through a pipe  \cite<e.g.>[]{rothlisberger1972,clarke2003}.
In the formulation here does not consider englacial storage of water, and thus that changes in hydraulic head at the top of the glacier are negligable in the evolution of the channel size. 
The evolution of subglacial channel size $S_g$ is given as
\begin{linenomath*}
  \begin{equation}
    \label{eq:dS_dt}
    \frac{\partial S_g}{\partial t} = C_2 \frac{Q \Delta h}{l} - C_2 (h_{o}-\frac{\Delta h}{2})^n\,S_g,
  \end{equation}
\end{linenomath*}
\noindent $C_1= (1-\rho_wc_pc_t)\,\frac{\rho_wg}{\rho_iL}$ and $C_2=2A(\frac{\rho_wg}{n})^n$ are constants (Table \ref{table:vpm}), $l$ is the length of subglacial channel, $Q$ is water discharge, $\Delta h$ is the hydraulic potential change from the glacier terminus to its top, $h_{o}= \frac{\rho_i}{\rho_w} h_{ice}$ is the ice overburden pressure, and $n$ is Glen's n.
The first term on the right side of the equation represents the opening of the channel through frictional heating, while the second term represents the creep closure of the channel from ice deformation.

Following the Darcy-Weisbach, head drop $\Delta h$ is,
\begin{linenomath*}
  \begin{equation}
    \label{eq:dh}
    \Delta h \,  = l \,\frac{1}{g}\,s\,f_r\,\frac{Q^2}{D_h^5}
  \end{equation}
\end{linenomath*}
\noindent where $f_r$ is a friction factor, $D_h$ is the hydraulic diameter, and $l$ is the length overwhich head change occurs. $s$ is a factor accounting for channel geometry \cite{hooke1990}, calculated as:
\begin{equation}
  \label{eq:Hf}
  s = \frac{2\,(\beta -\sin \beta)^2}{(\frac{\beta}{2}\,+\,\sin \frac{\beta}{2})^4},
\end{equation}
where $\beta$ is the central angle of the circular segment that comprises the channel (the so-called Hooke angle). Note that $\beta =\pi$ corresponds to a semi-circle and
smaller values of $\beta$ result in shallow, wide channels.

With knowledge of cross sectional area $S$, the shear stress, $\tau$, between the water and the channel bed is determined through the Darcy-Weisbach formulation
\begin{equation}
  \label{eq:tau}
  \tau_g=\frac{1}{8}\,f_r\,\rho_w\,v_g^2,
\end{equation}
% 
where  the water velocity $v_g = \frac{Q_w}{S_g}$.

The width of the flat channel floor $w_{cg}$, given angle $beta$ is 
\begin{equation}
  \label{eq:dh2wc}
  w_{cg} = 2  \sin \frac{\beta}{2} \sqrt{\frac{2\, S}{\beta -\sin \beta}}.
\end{equation}

@@

To parameterize the shear stress of water flowing across sediments in the subaerial channel,  the hydraulics parameterization presented in \citeA{tucker1997} is implemented.
Here, it is assumed that there is a conservation of mass and a sufficiently wide channel so that the hydraulic radius is consistent with flow depth, uniform flow, and the Darcy-Weisbach relationship.
The shear stress $\tau_f$ at the river bed is represented as
\begin{linenomath*}
  \begin{equation}
    \label{eq:DW_tau}
    \tau_f=\frac{\rho_w\,g^{\frac{2}{3}}\,f_p^{\frac{1}{3}}}{2}\, \Big(\frac{Q_w}{w_{cf}} \Big)^{\frac{2}{3}} \,\frac{\partial z_c}{\partial x}^{\frac{2}{3}},
  \end{equation}
\end{linenomath*}
where $\frac{\partial z_c}{\partial x}$ is the channel slope, and $f_p$ is the friction factor for subaerial systems.
Channel width $w_{cf}$ is 
\begin{equation}
  \label{eq:wcf}
  w_{cf} = k \, Q_w^e,
\end{equation}
% 
where $k$ is a constant and $e$ is an exponent commonly equal to $\frac{1}{2}$ \cite{leopold1953}.
Water velocity, $v_f$, is given by rearranging Equation \ref{eq:tau} as
\begin{equation}
  \label{eq:vf}
  v_f = \sqrt{\frac{8\,\tau_f}{f_p\,\rho_w}}.
\end{equation}
% 
As in Section~\ref{sect:sub_mode}, the width integrated shear stress is $\tau_{tf}=w_{cf}\,\tau_f$.

@@

The parameterizations above are applied to hydrological records from the Fieschergletscher in the Swiss Alps (info) and the Leverett glacier on the Greenland Ice Sheet (info).
The hydrographs applied to the subglacial and subaerial parameterizations represent the subglacial flow of water as it leaves the glacier and moves into a subaerial flow in front of the glacier, as water moves through the catchment.
Outputs of the parameterizations are meant to represent generalizable sediment transport characteristics from these hydrographs, rather than actual hydraulic conditions.

The formulation of the glacial system (Section~\ref{sect:sub_mode}) requires inputs of the ice thickness ($h_o$, Equation\,\ref{eq:dS_dt}), water discharge, $Q_w$,  Hooke angle (Equation~\ref{eq:Hf}; $\beta$), and friction factor (Equations~\ref{eq:dh}~and~\ref{eq:tau}; $f_r$). Values of these parameters are given in Table~\ref{table:vpm}.

@@

The parameterizations are first applied to a reference test case for each glacier that assumes a semi-circular subglacial channel ($\beta=\pi$) and friction factors $f_r$ and $f_p$ are tuned so that the model reproduces reasonable velocities for both  the Greenland Ice Sheet and alpine glaciers \cite<$\sim\,1.6\,$\unit{m}$^3$\,\unit{s}$^{-1}$>{werder2010b,chandler2013}.
Water velocity, shear-stress and width-integrated shear-stress in the subglacial and subaerial systems are presented below.

To characterize the variability in sediment discharge capacity in subglacial compared to subaerial systems further, the parameterizations are applied to range of different channel geometry factors ($\beta$ and $k$) and friction factors $f_r$ and $f_p$.
Model runs with parameter combinations  are culled, if their maximum subglacial water velocity exceeds $2\,$\unit{m}$^3$\,\unit{s}$^{-1}$ or if subaerial water velocity exceeds $1\,$\unit{m}$^3$\,\unit{s}$^{-1}$.
The standard deviation of water velocity, shear-stress and width-integrated shear-stress across the season long hydrographs are presented below.

\section{Results}

% there is more variability across the season subglacial and subaerial cases. peaks in subaerial cases lag peaks in subglacial cases. (Figure 2)

% For a fixed channel geometry, there is a single shear stress for a water discharge. In the subglacial case, peak water discharge that is 1.5 times the mean velocity. In Fieschercase, peak water discharge does not result in the greatest of any variable. In Leverett case the mean total shear stress can occur from 100 m3/s to 200 m3/s. The mean velocity and shear stress can occur across effectively all water discharges below 250 m3/s

% across a wide range of channel geometries
